Polynomial phase estimation by phase unwrapping

### 2013-02-03 paper.tex
moved stuff around.  Squashed stuff back into one papers.  Needs heay editing.

\title{Polynomial phase estimation by phase unwrapping 1: Identifiability and strong consistency}
%\title{Polynomial phase estimation by phase unwrapping}

@@

%A popular estimator is based on the \emph{discrete polynomial phase transform} (DPT) first introduced by Peleg and Porat \cite{Peleg_DPT_1995}.  The transform enables each parameter to be be estimated iteratively using the Fast Fourier transform in a way analogous to the frequency estimator of Rife and Boorstyn \cite{Rife1974, Quinn2001}.  The DPT estimator is characterized by good statistical performance and computational efficiency.  Variants of the DPT have been suggested by numerous authors including Golden and Friedlander \cite{Golden_mod_dpt_1998} and O'Shea \cite{Oshea_iterative_1996}. The recent \emph{high-order phase function} estimators are based on similar ideas \cite{Farquharson_another_poly_est_2005}.  A substantial drawback of all of these estimators is that they only work correctly over a small set of parameters, generally much smaller than the identifiable region.  This property appears to have been first noticed in \cite{McKilliam2009IndentifiabliltyAliasingPolyphase, McKilliam2009asilomar_polyest_lattice} and has been analysed further in \cite[Ch. 10]{McKilliam2010thesis}.
%
%Another estimator, described by Kitchen \cite{Kitchen_polyphase_unwrapping_1994}, is based on the phase unwrapping approach to frequency estimation suggested by Kay \cite{Kay1989}.  The primary advantage of Kitchen's estimator is that it requires only $O(N)$ arithmetic operations to compute. The primary disadvantage is that it only works correctly when the signal to noise ratio (SNR) is high. It was also observed in \cite{McKilliam2009asilomar_polyest_lattice} that Kitchen's estimator does not work correctly for all coefficients in the identifiable region, but, the problem is far less severe than for estimators based on the DPT \cite[Sec. 10.3]{McKilliam2010thesis}.

%An obvious estimator of the unknown coefficients is the least squares estimator (LSE).  This is also the maximum likelihood estimator (MLE) when the noise sequence $\{X_n\}$ is white and Gaussian.  When $m=0$ (phase estimation) or $m=1$ (frequency estimation) the LSE is an effective approach, being both computationally efficient and statistically accurate \cite{Quinn2009_dasp_phase_only_information_loss,Hannan1973,Quinn2001,McKilliam_mean_dir_est_sq_arc_length2010}\cite[Sec.~6.4~and~9.1]{McKilliam2010thesis}. When $m \geq 2$ the computational complexity of the LSE is large~\cite[Sec.~10.1]{McKilliam2010thesis}\cite{Abatzoglou_ml_chirp_1986}. For this reason many authors have considered alternative approaches to polynomial phase estimation. These can loosely be grouped into two classes, estimators based on polynomial phase transforms, such as the discrete polynomial phase transform (DPT)~\cite{Peleg_DPT_1995,Peleg1991_est_class_PPS_1991,Porat_asympt_HAF_DPT_1996} and the high order phase function~\cite{Farquharson_another_poly_est_2005}, and estimators based on phase unwrapping, such as Kitchen's unwrapping estimator~\cite{Kitchen_polyphase_unwrapping_1994}, the estimator of Djuric and Kay~\cite{Djuric_phase_unwrap_chirp_1990}, and Morelande's Bayesian unwrapping estimator~\cite{Morelande_bayes_unwrapping_2009_tsp}.

One estimation approach attempts to perform polynomial regression on the phase of the signal~\cite{Slocumb_polynomial_1994,Tretter1985,Kitchen_polyphase_unwrapping_1994,Djuric_phase_unwrap_chirp_1990,Morelande_bayes_unwrapping_2009_tsp}.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed.  In this two part series of papers we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping estimator} (LSU)~\cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009}\cite[Chap. 8]{McKilliam2010thesis}.  It was shown in \cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009} that the LSU estimator can be represented as a \emph{nearest lattice point problem}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In \cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder} \cite{Pohst_sphere_decoder_1981}, and \emph{Babai's nearest plane algorithm} \cite{Babai1986}. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for $N<50$ \cite{Jalden2005_sphere_decoding_complexity}.  Computing the Babai point requires only $O(N^2)$ operations, but its statistical performance is comparatively poor at low signal to noise ratio (SNR). A major point of interest is that the lattices considered are not \emph{random}, and therefore may admit fast nearest point algorithms. 
%This has been studied in~\cite[Sec.~4.2]{McKilliam2010thesis} where polynomial time algorithms where found that compute the nearest point exactly.  Unfortantely, although polynomial time, the algorithm are still computationally very slow in practice. 
% Two main questions were raised in \cite{McKilliam2009asilomar_polyest_lattice}:
% \begin{enumerate}
% \vspace{-0.1cm}
% \item What are the asymptotic statistical properties of the LSU estimator?
% \vspace{-0.18cm}
% \item Can the structure of the lattice be exploited to obtain fast (exact or approximate) nearest point algorithms and thereby a computationally efficient estimator?
% \vspace{-0.1cm}
% \end{enumerate}
% The purpose of this paper is to answer the first question. We prove that the LSU estimator is strongly consistent and asymptotically normally distributed. 
%We also apply another known approximate nearest point algorithm, called the $K$-best method \cite{Zhan2006_K_best_sphere_decoder}, that we show provides near sphere decoder performance, but can be computed in a reasonable amount of time if $N$ is less than about 300.
In this two part series of we derive the asymptotic statistical properties of the LSU estimator showing, under some assumptions on the noise $X_1,\dots,X_N$, that it is strongly consistent and asymptotically normally distributed.  Strong consistency is proved in this paper, while asymptotic normality is proved in the second part~\cite{McKilliam_pps2_2012}.  Similar results were stated without a complete proof in~\cite{McKilliam_polyphase_est_icassp_2011}.  Here, we give a proof.  The results here are also more general than in~\cite{McKilliam_polyphase_est_icassp_2011}, allowing for a wider class of noise distributions.

@@

It may be helpful to observe Figures~\ref{fig:circstatplot_zero} to~\ref{fig:circstatplot_cube}.  In these, the phase (divided by $2\pi$) of two distinct polynomial phase signals is plotted on the left, and on the right the principal component of the phase (also divided by $2\pi$) is plotted. The circles display the samples at the integers. Note that the samples of the principal components intersect.  The corresponding polynomial phase signals are aliases.

@@

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{plots/circstatfigzero-1.mps}
		\caption{The zeroth order polynomials $\tfrac{7}{10}$ (solid line) and $\tfrac{17}{10}$ (dashed line).}		
\label{fig:circstatplot_zero}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{plots/circstatfig-1.mps}
		\caption{The first order polynomials $\tfrac{1}{10}(3 + 8t)$ (solid) and $\tfrac{1}{10}(33 - 2t)$ (dashed line).}
		\label{fig:circstatplot_line}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{plots/circstatfigquad-1.mps}
		\caption{The quadratic polynomials $\tfrac{1}{10} (15 - 15 t + 4 t^2)$ (solid line) and $\tfrac{1}{10}(25 -  t^2)$  (dashed line).}
		\label{fig:circstatplot_quad}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{plots/circstatfigcube-1.mps}
		\caption{The cubic polynomials $\tfrac{1}{160} (174 + 85 t - 118 t^2 + 40 t^3)$ (solid line) and $\tfrac{1}{48} (84 + 19 t + 12 t^2 - 4 t^3)$  (dashed line).}
		\label{fig:circstatplot_cube}
\end{figure}

@@

\section{The least squares unwrapping estimator}\label{sec:least-squar-unwr}

@@

It is shown in~\cite[Sec~8.1]{McKilliam2010thesis}\cite{McKilliam2009asilomar_polyest_lattice} how this minimisation problem can be posed as that of computing a nearest lattice point in a particular lattice. Polynomial time algorithms that compute the nearest point are described in~\cite[Sec.~4.3]{McKilliam2010thesis}.  Although polynomial in complexity, these algorithms are not fast in practice.  The existence of practically fast nearest point algorithms for these lattices is an interesting open problem.  In this paper we focus on the asymptotic statistical properties of the LSU estimator, rather than computational aspects.

@@

The \emph{intrinsic mean} or \emph{Fr\'{e}chet mean} of $\Phi_n$ is defined as~\cite{McKilliam_mean_dir_est_sq_arc_length2010,Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005},
\begin{equation}\label{eq:intrmeandefn}
 \mu_{\text{intr}}  = \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2, 
% %&= \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})}\int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}\fracpart{\phi - \mu}^2f(\theta) d\theta. \label{eq:minunrappedmeandef}
\end{equation}
and the \emph{intrinsic variance} is
\[
\sigma_{\text{intr}}^2 = \expect\fracpart{\Phi_n - \mu_{\text{intr}}}^2 = \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2,
\]
where $\expect$ denotes the expected value.  Depending on the distribution of $\Phi_n$ the argument that minimises~\eqref{eq:intrmeandefn} may not be unique.  The set of minima is often called the~\emph{Fr\'{e}chet mean set}~\cite{Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005}.  If the minimiser is not unique we say that $\Phi_n$ has no intrinsic mean.  %We are now equipped to state the asymptotic properties of the LSU estimator.
We are now equipped to state the main result of this paper.

@@

\begin{theorem} \label{thm:asymp_proof}  (Strong consistency) 
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean and intrinsic variance $\sigma^2$, then $N^k \widehat{\lambda}_{k,N}$ converges almost surely to $0$ as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.
\end{theorem}

A proof of this theorem is given in the next section.  Proofs for the case when $m=0$ was given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and for the case when $m=1$ was given in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proof here takes a similar approach, but requires new techniques.  %Before giving the proof we make the following remarks.
The theorem gives conditions on the \emph{dealiased} difference $\dealias(\tilde{\mubf} - \widehat{\mubf})$ between the true coefficients $\tilde{\mubf}$ and the estimated coefficients $\widehat{\mubf}$ rather than directly on the difference $\tilde{\mubf} - \widehat{\mubf}$.   To see why this makes sense, consider the case when $m=0$, $\tilde{\mu}_0 = -0.5$ and $\widehat{\mu}_0 = 0.49$, so that $\tilde{\mu}_0 - \widehat{\mu}_0 = -0.99$.  However, the two phases are obviously close, since the phase $\pm 0.5$ are actually the same.  In this case 
\[
\dealias(\tilde{\mu}_0 - \widehat{\mu}_0) = \fracpart{\tilde{\mu}_0 - \widehat{\mu}_0} = 0.01.
\] 
The same reasoning holds for $m > 0$.

The requirement that $\Phi_1, \dots, \Phi_N$ be identically distributed will typically hold only when the complex random variables $X_1, \dots, X_N$ are identically distributed and circularly symmetric.  It would be possible to drop the assumption that $\Phi_1, \dots, \Phi_N$ be identically distributed, but this complicates the theorem statement and the proof.  In the interest of simplicity we only consider the case when $\Phi_1, \dots, \Phi_N$ are identically distributed here.  If $X_n$ is circularly symmetric with density function nonincreasing with magnitude $\abs{X_n}$, then, the corresponding $\Phi_n$ necessarily has zero intrinsic mean~\cite[Theorem~5.2,~page~78]{McKilliam2010thesis}.  Thus, our theorem covers commonly used distributions for $X_1, \dots, X_N$, such as the normal distribution.

Although we will not prove it here the assumption that $\Phi_1,\dots,\Phi_N$ have zero intrinsic mean is not only sufficient, but also necessary, for if $\Phi_1,\dots,\Phi_N$ have intrinsic mean $x \in [-\nicefrac{1}{2},\nicefrac{1}{2})$ with $x \neq 0$ then $\sfracpart{\widehat{\lambda}_{0,N} - x}\rightarrow 0$ almost surely as $N\rightarrow\infty$, and so $\widehat{\lambda}_{0,N}$ does not converge to zero.  On the other hand if $\Phi_1,\dots,\Phi_N$ do not have an intrinsic mean then $\widehat{\lambda}_{0,N}$ will not converge.

%The proof of asymptotic normality places requirements on the pdf $f$ of the phase noise.  The requirement that $\Phi_1, \dots, \Phi_N$ have zero intrinsic mean implies that $f(-\nicefrac{1}{2}) \leq 1$~\cite[Lemma~1]{McKilliam_mean_dir_est_sq_arc_length2010}, so the only case not handled is when equality holds, i.e., when $f(-\nicefrac{1}{2}) = 1$ or when $f(\fracpart{x})$ is discontinuous at $x = -\nicefrac{1}{2}$. In this exceptional case other expressions for the asymptotic variance can be found (similar to \cite[Theorem 3.1]{Hotz_circle_means_2011}), but this comes at a substantial increase in complexity and for this reason we have omitted them. %Distributions that do not satisfy these requirements are unlikely to be needed in practice.

@@

>>>>

%\title{Polynomial phase estimation by phase unwrapping 1: Identifiability and strong consistency}
\title{Polynomial phase estimation by phase unwrapping}

@@

An obvious estimator of the unknown coefficients is the least squares estimator (LSE).  This is also the maximum likelihood estimator (MLE) when the noise sequence $\{X_n\}$ is white and Gaussian.  When $m=0$ (phase estimation) or $m=1$ (frequency estimation) the LSE is an effective approach, being both computationally efficient and statistically accurate \cite{Quinn2009_dasp_phase_only_information_loss,Hannan1973,Quinn2001,McKilliam_mean_dir_est_sq_arc_length2010}\cite[Sec.~6.4~and~9.1]{McKilliam2010thesis}. When $m \geq 2$ the computational complexity of the LSE is large~\cite[Sec.~10.1]{McKilliam2010thesis}\cite{Abatzoglou_ml_chirp_1986}. For this reason many authors have considered alternative approaches to polynomial phase estimation. These can loosely be grouped into two classes, estimators based on polynomial phase transforms, such as the discrete polynomial phase transform (DPT)~\cite{Peleg_DPT_1995,Peleg1991_est_class_PPS_1991,Porat_asympt_HAF_DPT_1996} and the high order phase function~\cite{Farquharson_another_poly_est_2005}, and estimators based on phase unwrapping, such as Kitchen's unwrapping estimator~\cite{Kitchen_polyphase_unwrapping_1994}, the estimator of Djuric and Kay~\cite{Djuric_phase_unwrap_chirp_1990}, and Morelande's Bayesian unwrapping estimator~\cite{Morelande_bayes_unwrapping_2009_tsp}.

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping estimator} (LSU)~\cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009}\cite[Chap. 8]{McKilliam2010thesis}.  It was shown in \cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009} that the LSU estimator can be represented as a \emph{nearest lattice point problem}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In \cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder} \cite{Pohst_sphere_decoder_1981}, and \emph{Babai's nearest plane algorithm} \cite{Babai1986}. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for $N<50$ \cite{Jalden2005_sphere_decoding_complexity}.  Computing the Babai point requires only $O(N^2)$ operations, but its statistical performance is comparatively poor at low signal to noise ratio (SNR). A major point of interest is that the lattices considered are not \emph{random}, and therefore may admit fast nearest point algorithms. 
This has been studied in~\cite[Sec.~4.2]{McKilliam2010thesis} where polynomial time algorithms where found that compute the nearest point exactly.  Unfortantely, although polynomial time, the algorithm are still computationally very slow in practice. 
 Two main questions were raised in \cite{McKilliam2009asilomar_polyest_lattice}:
 \begin{enumerate}
 \vspace{-0.1cm}
 \item What are the asymptotic statistical properties of the LSU estimator?
 \vspace{-0.18cm}
 \item Can the structure of the lattice be exploited to obtain fast (exact or approximate) nearest point algorithms and thereby a computationally efficient estimator?
 \vspace{-0.1cm}
 \end{enumerate}
 The primary purpose of this paper is to answer the first question. Similar results were stated without a complete proof in~\cite{McKilliam_polyphase_est_icassp_2011}.  Here, we give a proof.  The results here are also more general than in~\cite{McKilliam_polyphase_est_icassp_2011}, allowing for a wider class of noise distributions.  We prove that the LSU estimator is strongly consistent and asymptotically normally distributed.  BLERG `small noise' assumption.

We also apply another known approximate nearest point algorithm, called the $K$-best method \cite{Zhan2006_K_best_sphere_decoder}, that we show provides near sphere decoder performance, but can be computed in a reasonable amount of time if $N$ is less than about 300. BLERG `better than everyting else and computationally reasonably for smallish $N \leq 300$'.

@@

It may be helpful to observe Figures~\ref{circstatplot_line} to~\ref{fig:circstatplot_cube}.  In these, the phase (divided by $2\pi$) of two distinct polynomial phase signals is plotted on the left, and on the right the principal component of the phase (also divided by $2\pi$) is plotted. The circles display the samples at the integers. Note that the samples of the principal components intersect.  The corresponding polynomial phase signals are aliases.

@@

% \begin{figure}[p]
% 	\centering
% 		\includegraphics[width=\linewidth]{plots/circstatfigzero-1.mps}
% 		\caption{The zeroth order polynomials $\tfrac{7}{10}$ (solid line) and $\tfrac{17}{10}$ (dashed line).}		
% \label{fig:circstatplot_zero}
% \end{figure}

\begin{figure}[t]
	\centering
		\includegraphics[width=\linewidth]{plots/circstatfig-1.mps}
		\caption{The first order polynomials $\tfrac{1}{10}(3 + 8t)$ (solid) and $\tfrac{1}{10}(33 - 2t)$ (dashed line).}
		\label{fig:circstatplot_line}
\end{figure}

% \begin{figure}[p]
% 	\centering
% 		\includegraphics[width=\linewidth]{plots/circstatfigquad-1.mps}
% 		\caption{The quadratic polynomials $\tfrac{1}{10} (15 - 15 t + 4 t^2)$ (solid line) and $\tfrac{1}{10}(25 -  t^2)$  (dashed line).}
% 		\label{fig:circstatplot_quad}
% \end{figure}

\begin{figure}[t]
	\centering
		\includegraphics[width=\linewidth]{plots/circstatfigcube-1.mps}
		\caption{The cubic polynomials $\tfrac{1}{160} (174 + 85 t - 118 t^2 + 40 t^3)$ (solid line) and $\tfrac{1}{48} (84 + 19 t + 12 t^2 - 4 t^3)$  (dashed line).}
		\label{fig:circstatplot_cube}
\end{figure}

@@

\section{The Least squares unwrapping estimator}\label{sec:least-squar-unwr}

@@

It is shown in~\cite{McKilliam2009asilomar_polyest_lattice} and \cite[Sec~8.1]{McKilliam2010thesis} how this minimisation problem can be posed as a nearest lattice point problem, and we briefly discuss this here.  Write $SS$ as
\[
SS(\mubf) = \sum_{n=1}^{N}{ \left( \Theta_{n} - W_n - \sum_{k = 0}^{m}{\mu_k n^k} \right)^2 },
\]
where $W_n = \round{\Theta_{n} - \sum_{k = 0}^{m}{\mu_k n^k} }$ for $n = 1, \dots, N$ are integers called \emph{wrapping variables}.  If we consider $W_1,\dots,W_N$ as nuisance parameters to be estimated then $SS$ can be written as a function of both $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$.  The least squares unwrapping estimator is then found by jointly minimising over $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$.   This joint minimisation problem can be solved by computing a nearest point in a lattice.  To see this write $SS$ as a function of both $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$ using vectors,
\begin{equation}\label{eq:SS_in_vector_form_with_u_nuisance}
SS(\mubf, \wbf) = \| \thetabf - \Xbf\mubf - \wbf \|^2,
\end{equation}
where $\|\cdot\|^2$ denotes the squared Euclidean norm, and $\Xbf$ is the rectangular Vandermonde matrix
\begin{equation}\label{eq:Xvandermonde}
\Xbf = \left[ \begin{array}{ccccc} 
1 & 1 & 1 & \cdots & 1 \\
1 & 2 & 4 & \cdots & 2^m \\
\vdots & \vdots  & \vdots & \ddots & \vdots \\
1 & N & N^2 & \cdots & N^m \\
\end{array} \right],
\end{equation}
and where the column vectors $\wbf = [W_1, \dots, W_N]^\prime$ and $\thetabf = [\Theta_1, \dots, \Theta_N]^\prime$.   For fixed $\wbf$ the minimiser of $SS(\mubf, \wbf)$ with respect to $\mubf$ is
\begin{equation} \label{eq:hat_theta_cal}
\mubf(\wbf) = (\Xbf^\prime\Xbf)^{-1}\Xbf^\prime( \thetabf - \wbf) = \Xbf^+( \thetabf - \wbf),
\end{equation}
where $\Xbf^+ = (\Xbf^\prime\Xbf)^{-1}\Xbf^\prime$ is the Moore-Penrose pseudoinverse of $\Xbf$.

We may thus compute $\widehat{\mubf}$ as $\widehat{\mubf} = \mu(\widehat{\wbf})$, where $\widehat{\wbf}$ is the minimiser of
\[
SS(\wbf) = SS(\mubf(\wbf),\wbf) = \| \Qbf\thetabf - \Qbf\wbf \|^2,
\]
where $\Qbf = \Ibf - \Xbf\Xbf^+$ and $\Ibf$ is the $N \times N$ identity matrix\footnote{We have slightly abused notation here by reusing $SS$. This should not cause any confusion as $SS(\mubf)$ and $SS(\mubf,\wbf)$ and $SS(\wbf)$ are easily told apart by their inputs.}.

Let $\Lambda$ be the lattice generated by $\Qbf$, that is $\Lambda = \{\Qbf\xbf \mid \xbf \in \ints^N \}$\footnote{The $N \times N$ projection matrix $\Qbf$ does not have full rank, it has rank $N-m-1$, and for this reason it is not strictly a generator matrix.  However, a generator can be constructed by removing $m+1$ consecutive columns from $\Qbf$. In our implementation the last $m+1$ columns are removed.}.  It follows that $\Qbf \wbf$ is a lattice point in $\Lambda$ and that minimising $SS(\wbf)$ is equivalent to finding the nearest lattice point in $\Lambda$ to $\Qbf\thetabf$.  Denote this point by $\Qbf\widehat{\wbf}$.  Now the estimate $\widehat{\mubf}$ is given by substituting $\widehat{\wbf}$ for $\wbf$ in~\eqref{eq:hat_theta_cal}.  After this procedure it is possible that the $\widehat{\mubf}$ obtained is not in the identifiable region but is instead an aliased version of the desired estimate. This can be resolved by computing $\dealias(\widehat{\mubf})$.  There exist algorithms that can compute a nearest point in $\Lambda$ that require a number of operations that is polynomial in $N$~\cite[Sec 4.3]{McKilliam2010thesis}.  Although polynomial in complexity, these algorithms are not fast in practice.  The existence of practically fast nearest point algorithms for these lattices is an interesting open problem.

@@

The \emph{intrinsic mean} or \emph{Fr\'{e}chet mean} of $\Phi_n$ is defined as~\cite{McKilliam_mean_dir_est_sq_arc_length2010,bwhk07a,Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005},
\begin{equation}\label{eq:intrmeandefn}
 \mu_{\text{intr}}  = \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2, 
% %&= \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})}\int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}\fracpart{\phi - \mu}^2f(\theta) d\theta. \label{eq:minunrappedmeandef}
\end{equation}
and the \emph{intrinsic variance} is
\[
\sigma_{\text{intr}}^2 = E\fracpart{\Theta - \mu_{\text{intr}}}^2 = \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2,
\]
where $\expect$ denotes the expected value.  Depending on the distribution of $\Phi_n$ the argument that minimises~\eqref{eq:intrmeandefn} may not be unique.  The set of minima is often called the~\emph{Fr\'{e}chet mean set}~\cite{Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005}.  If the minimiser is not unique we say that $\Phi_n$ has no intrinsic mean.  We are now equipped to state the asymptotic properties of the LSU estimator.

@@

\begin{theorem} \label{thm:asymp_proof} 
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean, intrinsic variance $\sigma^2$, and pdf $f$, then: 
\begin{enumerate}
\item (Strong consistency) $N^k \widehat{\lambda}_{k,N}$ converges almost surely to $0$ as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.
\item (Asymptotic normality) If $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and $f(-\nicefrac{1}{2}) < 1$ then the distribution of
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $(m+1)\times (m+1)$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k = \{0, 1, \dots, m\}$.
\end{enumerate}
\end{theorem}
The proof of Theorem~\ref{thm:asymp_proof} is contained within the next two sections. Section~\ref{sec:strongconstproof} proves strong consistency and Section~\ref{sec:centlimitproof} proves asymptotic normality.  The proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and for the case when $m=1$ were given in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proofs here take a similar approach, but require new techniques.  Before giving the proof we make the following remarks.

The theorem gives conditions on the \emph{dealiased} difference $\dealias(\tilde{\mubf} - \widehat{\mubf})$ between the true coefficients $\tilde{\mubf}$ and the estimated coefficients $\widehat{\mubf}$ rather than directly on the difference $\tilde{\mubf} - \widehat{\mubf}$.   To see why this makes sense, consider the case when $m=0$, $\tilde{\mu}_0 = -0.5$ and $\widehat{\mu}_0 = 0.49$, so that $\tilde{\mu}_0 - \widehat{\mu}_0 = -0.99$.  However, the two phases are obviously close, since the phase $\pm 0.5$ are actually the same.  In this case 
\[
\dealias(\tilde{\mu}_0 - \widehat{\mu}_0) = \fracpart{\tilde{\mu}_0 - \widehat{\mu}_0} = 0.01.
\] 
The same reasoning holds for $m > 0$.

The requirement that $\Phi_1, \dots, \Phi_N$ be identically distributed will typically hold only when the complex random variables $X_1, \dots, X_N$ are identically distributed and circularly symmetric.  It would be possible to drop the assumption that $\Phi_1, \dots, \Phi_N$ be identically distributed, but this complicates the theorem statement and the proof.  In the interest of simplicity we only consider the case when $\Phi_1, \dots, \Phi_N$ are identically distributed here.  If $X_n$ is circularly symmetric with density function nonincreasing with magnitude $\abs{X_n}$, then, the corresponding $\Phi_n$ necessarily has zero intrinsic mean~\cite[Theorem 5.2, page 78]{McKilliam2010thesis}.  Thus, our theorem covers commonly used distributions for $X_1, \dots, X_N$, such as the normal distribution.

The proof of asymptotic normality places requirements on the pdf $f$ of the phase noise.  The requirement that $\Phi_1, \dots, \Phi_N$ have zero intrinsic mean implies that $f(-\nicefrac{1}{2}) \leq 1$~\cite[Lemma~1]{McKilliam_mean_dir_est_sq_arc_length2010}, so the only case not handled is when equality holds, i.e., when $f(-\nicefrac{1}{2}) = 1$ or when $f(\fracpart{x})$ is discontinuous at $x = -\nicefrac{1}{2}$. In this exceptional case other expressions for the asymptotic variance can be found (similar to \cite[Theorem 3.1]{Hotz_circle_means_2011}), but this comes at a substantial increase in complexity and for this reason we have omitted them. %Distributions that do not satisfy these requirements are unlikely to be needed in practice.

@@

\section{Proof of asymptotic normality}\label{sec:centlimitproof}
